Ownership types in theory and practice (in Rust)

### 2023-01-23 slides/slides.tex
Added optional lifetime elision slides

\begin{frame}[fragile]{Borrowing: immutable references}
\begin{itemize}
  \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
  \item acquired using the \verb|&| operator
  \item unlimited amount of immutable references to same value are allowed (if no mutable references)

@@

  \item Will it compile? \pause No.
  % TODO: lifetime explanation
  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'
  \item Refactored version that compiles:\pause

@@

% TODO: will it compile: create multiple mutable references, but only use one
% However, the following code example \cite{rust-book} shows that it is important to note that the creation of multiple mutable and immutable references with scopes that do not overlap is allowed.

%   \begin{lstlisting}[language=Rust]
%   let mut s = String::from("hello");

%   let r1 = &s; // allowed
%   let r2 = &s; // allowed
%   println!("{} and {}", r1, r2);

%   let r3 = &mut s; // allowed
%   r3.push_str(" world");
%   println!("{}", r3);
%   \end{lstlisting}

% Since \emph{r1} and \emph{r2} are not used after line 5, that is where their scopes end.
% The scope of the mutable reference \emph{r3} only starts in line 7.
% Therefore, the scopes of the mutable reference and the two immutable references do not overlap, and the program compiles.

@@

>>>>

\begin{frame}[fragile]{Borrowing: immutable references}
\begin{itemize}
  \item read-only/shared references $\rightarrow$ compiler disallows modifying borrowed value
  \item \verb|&| and \verb|*| operators
  \item unlimited amount of immutable references to same value are allowed (if no mutable references)

@@

  \item Will it compile? \pause No.
  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
  \item Refactored version that compiles:\pause

@@

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s;
  let r2 = &s;
  println!("{} and {}", r1, r2);
  let r3 = &mut s;
  r3.push_str(" world");
  \end{lstlisting}

  \item Will it compile? \pause Yes.
  \item Why? \pause Lifetimes of immutable and mutable references don't overlap.

  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let r1 = &s; // lifetime 'a starts
  let r2 = &s; // lifetime 'b starts
  println!("{} and {}", r1, r2);
  // lifetimes 'a and 'b end
  let r3 = &mut s; // lifetime 'c starts
  r3.push_str(" world");
  \end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Borrowing: examples}
\begin{itemize}
  \begin{lstlisting}[language=Rust]
  let mut s = String::from("hello");
  let b = &mut a;
  Vec::push(&mut v, 12);
  println!("{}", *b);
  \end{lstlisting}
  \pause

  \begin{lstlisting}[language=Rust]
  let mut v = vec![10, 11];
  Vec::push(&mut v, 12); // lifetime 'b starts and ends
  let vptr = &mut v[1];  // lifetime 'a starts
  println!("{}", *vptr); // lifetime 'a ends
  \end{lstlisting}

  \item Will it compile? \pause No.
  % TODO: lifetime explanation
  \item Why not? \pause Borrowed value 'b' does not live as long as the reference 'a'.
  \item Refactored version that compiles:\pause

@@

\section{Extra slides}

\begin{frame}{Lifetime elision}
Here are the three rules:
\begin{itemize}
  \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
  \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
  \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
\end{itemize}
Otherwise, it is an error to elide an output lifetime.
\end{frame}

\begin{frame}[fragile]{Lifetime elision: examples}
\begin{lstlisting}[language=Rust]
fn print(s: &str); // elided
fn print<'a>(s: &'a str); // expanded

fn debug(lvl: u32, s: &str); // elided
fn debug<'a>(lvl: u32, s: &'a str); // expanded

fn substr(s: &str, until: u32) -> &str; // elided
fn substr<'a>(s: &'a str, until: u32) -> &'a str; // expanded

fn get_str() -> &str; // ILLEGAL, no inputs

fn frob(s: &str, t: &str) -> &str; // ILLEGAL, two inputs
fn frob<'a, 'b>(s: &'a str, t: &'b str) -> &str; // Expanded: Output lifetime is ambiguous
\end{lstlisting}
\end{frame}
